Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Updates

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Plots showing improvements in fit of the model over iterations
	on the Adult income dataset. For the expert simulation an LLM is used
	for the edge orientations. (a) Shows using total unexplained effect that is the
	sum of effect size between variables that don't have an edge between them. (b)
	Shows the log-likelihood of the overall model.
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

Our procedure closely resembles certain score-based automated causal discovery
methods, such as Greedy Equivalence Search (GES), where the algorithm
iteratively adds or removes edges that improve the scoring metric the most.
Similarly, a greedy version of our approach would involve adding an edge
between the pair of variables with the highest unexplained correlation. While
both methods aim to find modifications that most improve the model at each
step, a key distinction lies in the nature of the evaluation criteria: unlike
standard scoring metrics, our measure of association is not decomposable. A key
property of scoring metrics is that they should be decomposible, i.e., they can
be expressed as the sum of scores of nodes given its parents. This means adding
or removing an edge affects only a localized part of the model. In contrast,
our measure of association is global, meaning that modifying an edge in one
part of the model can influence association values elsewhere.

Given a DAG $ G = (V, E) $, the total unexplained correlation, $\tau$ in define as:
\begin{equation}
	\tau = \sum_{\{ (X, Y) \in V | X \rightarrow Y \not \in E \wedge Y \rightarrow X \not \in E \} } \phi(X, Y, \mathrm{pa}_G(X) \cup \mathrm{pa}_G(Y)))
\end{equation}

>>>>

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{Plots showing improvements in fit of the model over iterations
	on the Adult income dataset. For the expert simulation an LLM is used
	for the edge orientations. (a) Shows using total residual association that is the
	sum of effect size between variables that don't have an edge between them. (b)
	Shows the log-likelihood of the overall model.
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

Our procedure closely resembles certain score-based automated causal discovery
methods, such as Greedy Equivalence Search (GES), where the algorithm
iteratively adds or removes edges that improve the scoring metric the most.
Similarly, a greedy version of our approach would involve adding an edge
between the pair of variables with the highest residual association. While
both methods aim to find modifications that most improve the model at each
step, a key distinction lies in the nature of the evaluation criteria: unlike
standard scoring metrics, our measure of association is not decomposable. A key
property of scoring metrics is that they should be decomposible, i.e., they can
be expressed as the sum of scores of nodes given its parents. This means adding
or removing an edge affects only a localized part of the model. In contrast,
our measure of association is global, meaning that modifying an edge in one
part of the model can influence association values elsewhere.

Given a DAG $ G = (V, E) $, the total residual association, $\tau$ in define as:
\begin{equation}
	\tau = \sum_{\substack{X, Y \in V \\ X \rightarrow Y, Y \rightarrow X \not \in E}}   \phi(X, Y, \mathrm{pa}_G(X) \cup \mathrm{pa}_G(Y))
\end{equation}
